\textbf{Beasts of Prey or Rational Animals?\\ Private Governance in Brazil's \textit{Jogo do Bicho}}

### 2018-10-16 paper-bicho/bicho.tex
add text sec 2

The threat of violence is balanced with financial benefits to low-rank members of the \textit{jogo do bicho} organisation. \emph{Bicheiros} keep all tips they receive from players, often have small expenses covered by their bosses, and may request interest-free loans to pay for healthcare treatment or other unexpected bills \citep{labronici2012paratodos}.

@@

Evolutionary game theory \citep{axelrod1984evolution, axelrod1985achieving, smith1982evolution} and experimental studies \citep{isaac1984divergent, marwell1981economists} have both demonstrated that long-term cooperation is possible whenever players expect future pay-offs to be higher than present ones. Fear of retaliation induces individuals not to cheat. Nevertheless, illegal organisations tend to discount the future more heavily than the other groups, what makes cooperative behaviour uncommon in criminal groups \citep{skarbek2011governance,skarbek2012prison,skarbek2014social}.

The \emph{jogo do bicho} is an exception to this rule. Here I analyse two means by which the \textit{bicheiros} elicit voluntary cooperation from gamblers and members of the community, costly signalling and reputation strategies. In the following section I discuss how the \textit{jogo} colludes with corrupt politicians and civil servants for mutual benefit.

>>>>

Despite these serious events, killings are rare in the animal game. Since the lottery bosses can credibly indicate that violence is a low-cost option for them, the mere threat of punishment is enough to induce cooperation. This is a good strategy for the \textit{banqueiros}. Interestingly, the fact that they had committed violent crimes in the past reduces the need to commit them in the present, as and a result they can spend less money on security and increase profits. As it happens in many traditional markets, if a group is able to form a cartel, they can increase the price of their services without fearing immediate competition. The same logic is valid for the \textit{jogo do bicho}, although the means it employs are unconventional.

The threat of violence is not the only means the \textit{bicheiros} have at hand. They balance it  with financial benefits to low-rank members of the organisation. For instance, street \emph{bicheiros} keep all tips they receive from players, often have small expenses covered by their bosses, and may request interest-free loans to pay for healthcare treatment or other unexpected bills \citep{labronici2012paratodos}.

@@

Evolutionary game theory \citep{axelrod1984evolution, axelrod1985achieving, smith1982evolution} and empirical studies \citep{isaac1984divergent, ostrom1990governing} have both demonstrated that long-term cooperation is possible in a great number of situations. The main requirement for sustained cooperation is that players should believe that future pay-offs will be higher than present ones. If that condition is true, fear of retaliation will induce individuals not to cheat.

In theory, the same should also apply to illegal organisations. Yet in practice we see that criminal groups tend to be short-term oriented, or, in more technical terms, they tend to discount the future more heavily that most people. This makes cooperative behaviour rather uncommon among criminal organisations, and there is considerable evidence suggesting that illegal groups have considerable difficulty to maintain themselves in the long run.

The \emph{jogo do bicho} is an exception to this rule. The game has been running for more than a century with no considerable interruption and with moderately low levels of violence -- at least when compared to other illegal activities such as drug trafficking. Moreover,

 Here I analyse two means by which the \textit{bicheiros} elicit voluntary cooperation from gamblers and members of the community, costly signalling and reputation strategies. In the following section I discuss how the \textit{jogo} colludes with corrupt politicians and civil servants for mutual benefit.
